# Basin characterisation {#sec-basin-characterisation}

```{r}
#| include: false

```

# Theory {#sec-basin-characterisation-theory .unnumbered}

In this chapter, we look at **land cover** and the definition of **hydrological response units (HRU's)** and their importance in hydrological modeling.

## Land use and land cover

Coming soon...

## Hydrological response units (HRUs)

**Hydrological response units (HRUs)** are areas that have common properties which are important from the hydrological perspective for runoff generation: for instance similar slope, elevation, aspect, soil type, vegetation cover, landuse, etc.

Comig soon...

# Exercise 2: Basin characterisation {#sec-basin-characterisation-exercise .unnumbered}

The goal of the second exercise is to fill in @tbl-example-basin-characterization with data from your catchment. This table includes key characterisations relevant to hydrological modelling for your specific catchment area.

In this exercise, we will also learn how to create a map visualizing your study area and a map illustrating the different land cover types, similar to the one shown in @fig-studyarea-dem and @fig-studyarea-landcover for the upper ZRB.

![Overview over the upper Zarafshan River Basin. The digital elevation model is from @srtm_2020, the glacier outlines (clear blue) are from the glacier Randolph Inventory (@RGI_Consortium_2023)](/images/Climate_change_impact_study/Basin_characterisation/Studyarea_DEM.pdf){#fig-studyarea-dem width="300"}

![Land cover map of the upstream part of the Zarafshan river basin from the ESA WorlCover dataset from @zanaga_2021_5571936.](/images/Climate_change_impact_study/Basin_characterisation/Studyarea_Landcover.pdf){#fig-studyarea-landcover width="300"}

This will be done in several parts, each supported by tutorial videos in both English and Russian:

-   Part 1: DEM Land Cover

-   Part 2: Glacier outlines and Study area map

-   Part 3: HRU's

-   Part 4: Discharge data

-   Part 5: Climate data

## Part 1: DEM and Land Cover {#sec-basin-characterisation-exercise-Part1 .unnumbered}

In this chapter, we will fill out the sections Geography and Land Cover in Table 1 and download the digital elevation model and land cover data within our catchment boundaries, which we delineated in @sec-watershed-delineation. To download the DEM and Land cover data we will use the Google Earth Engine (GEE). Google Earth Engine is a cloud-based platform for planetary-scale environmental data analysis. It hosts a vast amounts of environmental data, enabling users to download and perform various analyses and processing tasks.

In this part of the exercise, we will use the GEE to download the SRTM Digital Elevation Model and Land Cover data (Source) and calculate Land cover type statistics using the [GEE Link](https://code.earthengine.google.com/c8b1300f5d547c96a5892c723e5f2532) in combination with the following tutorial.

There are several other ways to download the DEM and Land Cover data. Here is a turorial how to donwload the DEM from the[Earth Explorer](https://hydrosolutions.github.io/caham_book/appendix_c_quick_guides.html#section-earth-explorer-download-srtm) or use the [QGIS Plugin](https://hydrosolutions.github.io/caham_book/appendix_c_quick_guides.html#sec-srtm-plugin). Land cover data can also be obtained directly from the [ESA WorldCover webpage](https://esa-worldcover.org/en/data-access).

**English version:**

```{r}
#| echo: false
library(vembedr)
embed_url("https://www.youtube.com/watch?v=fGmNkA1_lcg") %>%
  use_align("center")
```

**Russian version:**

```{r}
#| echo: false
library(vembedr)
embed_url("https://www.youtube.com/watch?v=dn3OhQAr4a0") %>%
  use_align("center")
```

## Part 2: Glacier outlines and Study area map {#sec-basin-characterisation-exercise-Part2 .unnumbered}

## Part 3: Hydrological response units (HRUs) {#sec-basin-characterisation-exercise-Part3 .unnumbered}

## Part 4: Discharge data {#sec-basin-characterisation-exercise-Part4 .unnumbered}

## Part 5: Climate data {.unnumbered}

## References {#sec-basin-characterisation-references}
